\textbf{Urban mobility in Latin America after COVID-19}

### 2024-01-20 manuscript/manuscript.qmd
update intro

The COVID-19 pandemic resulted in a notable reduction in mobility within cities. Nonpharmaceutical interventions to contain the spread of COVID‐19 reduced overall levels of mobility. Especially during lockdowns, mobility recorded reductions in the frequency, distance and time of trips in cities across the globe. Rises in teleworking, online schooling and remote shopping activity reduced the need to travel for work, education, shopping and leisure. Coupled with fears of crowded public spaces, nonpharmaceutical interventions prompted more geographically localised mobility patterns. Reductions in mobility were however socioeconomically highly unequal, with more affluent households recording greater drops in mobility levels. Wealthier individuals can afford to move less as they are more likely to work on jobs that can be done remotely, while individuals from more deprived background tend to engage in face-to-face activities requiring physical presence.

* Context and importance of human mobility: public health , climate change\
* The pandemic resulted in major changes in human mobility - unequal impacts across socioeconomic groups\
* Gap 1: Little work on understanding the long-term changes in mobility - recovery\
* Gap 2: little work on less developed countries - where inequalities are more pronounced\
* Aim - focus on South American countries as case study.

>>>>

The COVID-19 pandemic resulted in a notable reduction in mobility within cities.
Nonpharmaceutical interventions to contain the spread of COVID‐19 reduced overall levels of mobility.
Especially during lockdowns, mobility recorded reductions in the frequency, distance and time of trips in cities across the globe.
Rises in teleworking, online schooling and remote shopping activity reduced the need to travel for work, education, shopping and leisure.
Coupled with fears of crowded public spaces, nonpharmaceutical interventions prompted more geographically localised mobility patterns.

However, reductions in mobility levels were highly unequal reflecting existing socioeconomic inequalities.
In most countries, affluent individuals tended to record the greatest drops in mobility levels as they are predominantly employed in knowledge-intensive jobs which can be done fully or partly remotely.
During the COVID-19 pandemic, the adoption of remote work reduced the need of commuting for knowledge-intensive, non-public facing jobs.
At the same time, individuals from less privileged socioeconomic backgrounds displayed less pronounced declines mirroring the nature of their work requiring public-facing, face-to-face interaction, and thus a requirement for daily work commutes.

Thus, while a growing body of empirical evidence has contributed to advancing our understanding of the impacts of the COVID-19 pandemic on spatial mobility within cities, existing research has focused on more developed countries and the immediate effects of the pandemic during 2020.
Less is known about the longer term patterns of resilience in urban mobility in less developed countries.
Assessing the extent to which the intensity of movement has returned back to pre-pandemic levels across socioeconomic groups is important to understand the potentially unequal impacts of hybrid working in the society.

A key barrier to continously monitor changes in geographic mobility patterns in less developed countries during and post the COVID-19 pandemic has been the lack of suitable data.
Traditionally census and survey data have been employed to analyse human mobility patterns in these countries.
Yet, these data streams are not frequently updated and suffer from slow releases, with census data for example
being collected over intervals of ten years in most countries.
These data streams thus lack the temporal granularity to analyse population movements over short-time periods.
Data resulting from social interactions on digital platforms has emerged as an unique source of information to capture human population movement at scale in less developed countries.
Particularly location data from mobile phone applications have become a prominent source to sense patterns of human mobility at higher geographical and temporal resolution over in real time.

We aim to

-   Context and importance of human mobility: public health , climate change\
-   The pandemic resulted in major changes in human mobility - unequal impacts across socioeconomic groups\
-   Gap 1: Little work on understanding the long-term changes in mobility - recovery\
-   Gap 2: little work on less developed countries - where inequalities are more pronounced\
-   Aim - focus on South American countries as case study.

### 2024-01-20 manuscript/manuscript.tex
update intro

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=2.54cm,right=2.54cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{lipsum} \usepackage{libertine}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Urban mobility in Latin America after COVID-19},
  pdfauthor={Carmen Cabrera-Arnau; Francisco Rowe; Miguel González-Leonardo; Andrea Nasuto; Ruth Neville},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
The COVID‐19 pandemic has impacted the national systems of population
movement around the world.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

The COVID-19 pandemic resulted in a notable reduction in mobility within
cities. Nonpharmaceutical interventions to contain the spread of
COVID‐19 reduced overall levels of mobility. Especially during
lockdowns, mobility recorded reductions in the frequency, distance and
time of trips in cities across the globe. Rises in teleworking, online
schooling and remote shopping activity reduced the need to travel for
work, education, shopping and leisure. Coupled with fears of crowded
public spaces, nonpharmaceutical interventions prompted more
geographically localised mobility patterns. Reductions in mobility were
however socioeconomically highly unequal, with more affluent households
recording greater drops in mobility levels. Wealthier individuals can
afford to move less as they are more likely to work on jobs that can be
done remotely, while individuals from more deprived background tend to
engage in face-to-face activities requiring physical presence.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=2.54cm,right=2.54cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{lipsum} \usepackage{libertine}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Urban mobility in Latin America after COVID-19},
  pdfauthor={Carmen Cabrera-Arnau; Francisco Rowe; Miguel González-Leonardo; Andrea Nasuto; Ruth Neville},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
The COVID‐19 pandemic has impacted the national systems of population
movement around the world.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced]}{\end{tcolorbox}}\fi

@@

The COVID-19 pandemic resulted in a notable reduction in mobility within
cities. Nonpharmaceutical interventions to contain the spread of
COVID‐19 reduced overall levels of mobility. Especially during
lockdowns, mobility recorded reductions in the frequency, distance and
time of trips in cities across the globe. Rises in teleworking, online
schooling and remote shopping activity reduced the need to travel for
work, education, shopping and leisure. Coupled with fears of crowded
public spaces, nonpharmaceutical interventions prompted more
geographically localised mobility patterns.

However, reductions in mobility levels were highly unequal reflecting
existing socioeconomic inequalities. In most countries, affluent
individuals tended to record the greatest drops in mobility levels as
they are predominantly employed in knowledge-intensive jobs which can be
done fully or partly remotely. During the COVID-19 pandemic, the
adoption of remote work reduced the need of commuting for
knowledge-intensive, non-public facing jobs. At the same time,
individuals from less privileged socioeconomic backgrounds displayed
less pronounced declines mirroring the nature of their work requiring
public-facing, face-to-face interaction, and thus a requirement for
daily work commutes.

Thus, while a growing body of empirical evidence has contributed to
advancing our understanding of the impacts of the COVID-19 pandemic on
spatial mobility within cities, existing research has focused on more
developed countries and the immediate effects of the pandemic during
2020. Less is known about the longer term patterns of resilience in
urban mobility in less developed countries. Assessing the extent to
which the intensity of movement has returned back to pre-pandemic levels
across socioeconomic groups is important to understand the potentially
unequal impacts of hybrid working in the society.

A key barrier to continously monitor changes in geographic mobility
patterns in less developed countries during and post the COVID-19
pandemic has been the lack of suitable data. Traditionally census and
survey data have been employed to analyse human mobility patterns in
these countries. Yet, these data streams are not frequently updated and
suffer from slow releases, with census data for example being collected
over intervals of ten years in most countries. These data streams thus
lack the temporal granularity to analyse population movements over
short-time periods. Data resulting from social interactions on digital
platforms has emerged as an unique source of information to capture
human population movement at scale in less developed countries.
Particularly location data from mobile phone applications have become a
prominent source to sense patterns of human mobility at higher
geographical and temporal resolution over in real time.

We aim to
